---
title: "HW #6 – Economic Development"
author: "Caleb C"
date: "10/22/2025"
output:
  pdf_document: 
    extra_dependencies: ["mathtools", "amsmath", "amsfonts"]
  html_document:
    df_print: paged
subtitle: "Select Questions from Chapter 7 and 8"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
library(dplyr)
library(tibble)
library(ggplot2)
```

# Chapter 7 – Measuring Productivity

### Question 5 – Role of Factor Accumulation vs Productivity in Cross‑Country Income Gaps

**Method and Formulae**  
To decompose cross-country differences in output per worker, we assume a Cobb–Douglas production function in per-worker terms  

\[
y \;=\; A \, k^{\alpha} \, h^{1-\alpha},
\]

where \(y\) denotes output per worker, \(k\) physical capital per worker, \(h\) human capital per worker, and \(A\) total factor productivity (TFP).  The standard capital share of income is set to \(\alpha = \tfrac13\).  All variables are expressed relative to the United States, which is normalized to 1.  

Under this specification:

* **Factor accumulation** is captured by  
  \[
  F \;=\; k^{\alpha}\,h^{1-\alpha},
  \]
  which aggregates the contributions of physical and human capital.

* **Relative productivity** is measured by  
  \[
  A \;=\;\frac{y}{F},
  \]
  isolating the portion of income differences that cannot be explained by factor inputs alone.  

A logarithmic decomposition of \(\ln y = \ln F + \ln A\) allows us to express the percentage contribution of inputs and productivity to each country’s income gap vis-à-vis the United States.

**Calculation**
\vspace{-10pt}
```{r q5_data}
alpha <- 1/3          # capital share 

countries <- tribble(             # input given data 
  ~Country,      ~y,   ~k,   ~h,
  "Netherlands", 0.78, 1.12, 0.90,
  "Paraguay",    0.16, 0.15, 0.70,
  "Pakistan",    0.15, 0.08, 0.47
) %>% 
  mutate(                                   # desired calculations 
    Factor_Accum = k^alpha * h^(1 - alpha),
    Productivity = y / Factor_Accum,
    Factor_Share = 100 * log(Factor_Accum) / log(y),
    Prod_Share   = 100 * log(Productivity) / log(y)
  )
```

```{r q5_table, echo=FALSE}
knitr::kable(             # table output 
  countries %>%
    select(Country, y, k, h,
           `Factor Accum.` = Factor_Accum,
           Productivity,
           `Factor Share (%)`  = Factor_Share,
           `Prod. Share (%)`   = Prod_Share),
  digits = 3,
  caption = "Question 5 – Decomposing the 2017 income gap relative to the United States"
)
```

**Interpretation**  
Pakistan’s gap is 71 % factor‑driven, whereas the Netherlands’ gap is 87 % TFP‑driven; Paraguay lies between, with roughly a 50‑50 split. Therefore, factor accumulation plays the largest role in explaining income relative to the United States in Pakistan; productivity plays the largest role in explaining income relative to the United States in the Netherlands.

---

### Question 6 – Growth Accounting for Argentina, Uruguay & Panama (1975‑2017)

**Method and Formulae**  
We assume a Cobb–Douglas production function in per-worker terms,  

\[
y = A\,k^{\alpha}h^{1-\alpha},
\]

with a typical capital share \(\alpha \approx \tfrac13\).

Log-differentiating yields  

\[
\hat{y} \;=\; \hat{A} \;+\; \alpha\,\hat{k} \;+\; (1-\alpha)\,\hat{h},
\]

where

* \(\hat{y}\) = growth of output per worker,  
* \(\hat{k}\) = growth of physical capital per worker,  
* \(\hat{h}\) = growth of human capital per worker,  
* \(\hat{A}\) = total factor productivity (TFP) growth (the Solow residual),  

so that we can derive

* **TFP growth:** \(\hat{A} = \hat{y} - \bigl[\alpha\,\hat{k} + (1-\alpha)\,\hat{h}\bigr]\)  
* **Factor accumulation growth:** \(\hat{FA} = \alpha\,\hat{k} + (1-\alpha)\,\hat{h}\).

**Calculation**
```{r q6_data}
tbl <- data.frame(                                 # Input given data
  Country   = c("Argentina", "Uruguay", "Panama"),
  g_y       = c(0.31, 1.46, 1.59),   # Growth of output per worker (%)
  g_k       = c(0.88, 1.59, 1.97),   # Growth of capital per worker (%)
  g_h       = c(0.79, 0.73, 0.84)    # Growth of human capital per worker (%)
)

alpha <- 1/3          # capital share

tbl <- tbl |>
  mutate(             # desired calculations
    g_FA = alpha * g_k + (1 - alpha) * g_h, # Factor accumulaton growth
    g_A = g_y - g_FA,                       # TFP growth
    share_FA = 100 * g_FA / g_y, # % of growth due to FA
    share_A = 100 * g_A / g_y # % of growth due to TFP
)
```

```{r q6_table, echo=FALSE}
knitr::kable(
  tbl |>
    select(Country, g_y, g_k, g_h, g_FA, g_A, share_FA, share_A),
    digits = 2,
    col.names = c("Country",
                  "\\protect$\\hat{y}$ (\\%)", 
                  "\\protect$\\hat{k}$ (\\%)", 
                  "\\protect$\\hat{h}$ (\\%)", 
                  "\\protect$\\hat{FA}$ (\\%)", 
                  "\\protect$\\hat{A}$ (\\%)", 
                  "FA Share of Growth (\\%)", 
                  "TFP Share of Growth (\\%)"),
    caption = "Question 6 – Growth rate decomposition, 1975‑2017",
  format = "latex",
  booktabs = TRUE,
  escape = FALSE
)
```

**Interpretation**  
In Argentina, factor accumulation accounts for the largest share of growth contribution among the three countries at 264.52% of growth; in Uruguay, total factor productivity accounts for the largest share of growth contribution among the three countries at 30.37%.

---

# Chapter 8: The Role of Technology in Growth

### Question 1 – The Rivalry and Excludability of Goods

| Good | Rival? | Excludable? |
|------|--------|-------------|
| National defense | No | No |
| A cookie | Yes | Yes |
| Password‑protected website | No | Yes |
| Fruit on a public square tree | Yes | No |

---

### Question 2 – Evaluating Patent Protection for Life Saving Drugs

**Advantages of patents**  

* Monopoly rents raise expected returns and stimulate risky, long-term R&D.  
* Patent disclosure places technical details in the public domain once exclusivity ends.  
* Clear intellectual property rights facilitate venture financing, licensing, and strategic alliances.  
* Temporary exclusivity aligns with regulatory approval timelines, helping recoup clinical-trial costs.  


**Disadvantages of patents**  

* Deadweight loss from monopoly pricing; patients in low income groups may be priced out.  
* Follow-on innovation can be blocked by “patent thickets” and evergreening strategies.  
* High prices distort R&D toward diseases with lucrative markets, neglecting rare or tropical illnesses.  
* Administrative, monitoring, and litigation costs consume public and private resources.  

**Possible alternative:** *Prize funds* (or advance market commitments) pay inventors for reaching pre‑specified health targets, after which the technology enters the public domain.  
*Advantages:* keeps the incentive to innovate while allowing competitive generic supply immediately; can be targeted to neglected diseases; reduces IP-related litigation.  
*Disadvantages:* requires large upfront public financing and careful contract design; risks mispricing the reward; demands robust verification of performance; may dampen incremental proprietary improvements without supplementary incentives.  

---

### Question 3 – Temporary Increase in \(\gamma_A\) in the One‑Country Model

```{r q3_setup, message=FALSE, include=FALSE}
library(tidyverse)

# Parameters
T          <- 60                 # periods
gamma_base <- 0.05               # baseline R&D share
gamma_high <- 0.10               # temporary high R&D share
t_start    <- 10                 # shock starts
t_end      <- 20                 # shock ends
mu         <- 10                 # idea price
A0         <- 1                  # initial productivity

# γ_A path
gammaA <- rep(gamma_base, T)
gammaA[(t_start+1):(t_end)] <- gamma_high      # R uses 1-indexing

# Growth rates of A
gA <- gammaA/mu                 #  \hat A_t

# Simulate paths
A  <- numeric(T); y  <- numeric(T)
A[1] <- A0; y[1] <- A0*(1-gammaA[1])

for(t in 2:T){
  A[t] <- A[t-1]*(1+gA[t-1])
  y[t] <- A[t]*(1-gammaA[t])
}

baseline_A <- cumprod(c(A0, rep(1+gamma_base/mu, T-1)))[1:T]
baseline_y <- baseline_A*(1-gamma_base)

sim_df <- tibble(
  period = 0:(T-1),
  A_shock = A,
  A_base  = baseline_A,
  y_shock = y,
  y_base  = baseline_y
) %>% pivot_longer(-period,
                   names_to = c("series","scenario"),
                   names_sep = "_",
                   values_to = "value")
```

```{r q3_graphs, echo=FALSE}
ggplot(sim_df %>% filter(series=="A"),
aes(period, value, colour = scenario, linetype = scenario)) +
geom_line(size = 1) +
labs(y = "Productivity A (ratio scale)",
x = "Time",
title = "Path of Productivity A") +
scale_y_log10() +
theme_minimal()

ggplot(sim_df %>% filter(series=="y"),
aes(period, value, colour = scenario, linetype = scenario)) +
geom_line(size = 1) +
labs(y = "Output per worker y (ratio scale)",
x = "Time",
title = "Path of Output per Worker y") +
scale_y_log10() +
theme_minimal()
```

---

### Question 4 – Quantifying the ‘Catch‑Up’ Time after Raising \(\gamma_A\)

Parameters: \(L=1,\;\mu=5\).  Initially \(\gamma_A=0.5\) then *permanently* raised to \(0.75\).

```{r q4_numeric}
mu        <- 5
gamma_old <- 0.50
gamma_new <- 0.75

# (i) Growth rates of productivity and output per worker
g_A_old <- mu * gamma_old        # 2.5 % per year
g_A_new <- mu * gamma_new        # 3.75 % per year

# (ii) Instantaneous output drop (smaller production labor share)
drop <- (1 - gamma_new) / (1 - gamma_old)  # 50 % fall

# (iii) Time until the new path catches up with the old path
delta_g <- g_A_new - g_A_old      # 1.25 pp faster
time_yrs <- log(1 / drop) / delta_g   # ln(2)/1.25 approx. 0.55 years
round(c(g_A_old = g_A_old, g_A_new = g_A_new, catch_up_years = time_yrs), 3)
```

Hence **output initially halves** but, because productivity now grows 1.25 pp faster, the economy regains the lost ground in **about 0.55 years (about 7 months)** and thereafter stays ahead.

---

### Question 5 – Two‑Country Model with Cost of Copying

The follower’s *copying cost* is \(\mu_c = \mu_i\,(A_1/A_2)^{-\beta}\) with \(0<\beta<1\), equal labor forces, and technology‑creation intensity \(\gamma_{A,i}\).  At steady state the two technology growth rates are equal:

\[
\mu_i \gamma_{A,1} = \mu_i \gamma_{A,2} + \mu_i\,(A_1/A_2)^{-\beta} 
\quad\Rightarrow\quad
\bigl(\tfrac{A_1}{A_2}\bigr)^{-\beta} = \gamma_{A,1}-\gamma_{A,2}.
\]

Therefore the **steady‑state technology ratio** is

\[
\boxed{\displaystyle \frac{A_1}{A_2} = (\gamma_{A,1}-\gamma_{A,2})^{-1/\beta}}.
\]

*Comparative statics:* A *smaller* \(\beta\) (flatter copying‑cost curve) implies a **larger** steady‑state gap because copying becomes cheaper when the gap is wide.

#### Numerical Example

Parameters: \(\beta=1/2,\;\mu_i=10,\;\gamma_{A,1}=0.2,\;\gamma_{A,2}=0.1\).

```{r q5_numeric}
beta <- 0.5
g1   <- 0.2
g2   <- 0.1
A_ratio <- (g1 - g2)^(-1/beta)  # = 100
A_ratio
```

Country 1 ends up with **A_1/A_2 = 100**, the leader’s technology level is two orders of magnitude higher than the follower’s.

---

*End of Assignment*