---
title: "HW #6 – ECON 4100: Economic Development"
author: "Caleb Cayouette"
date: "10/22/2025"
output:
  pdf_document:
    extra_dependencies: ["mathtools", "amsmath", "amsfonts", "float"]
  html_document:
    df_print: paged
subtitle: "Select Questions from Ashraf & Weil (2025): Chapters 7 and 8"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
library(dplyr)
library(tibble)
library(ggplot2)
library(kableExtra)
library(latex2exp)
library(grid)
```

# Chapter 7 – Measuring Productivity

### Question 5 – Role of Factor Accumulation vs Productivity in Cross‑Country Income Gaps

**Method and Formulae**  
To decompose cross-country differences in output per worker, we assume a Cobb–Douglas production function in per-worker terms  

\[
y \;=\; A \, k^{\alpha} \, h^{1-\alpha},
\]

where \(y\) denotes output per worker, \(k\) physical capital per worker, \(h\) human capital per worker, and \(A\) total factor productivity (TFP).  The standard capital share of income is set to \(\alpha = \tfrac13\).  All variables are expressed relative to the United States, which is normalized to 1.  

Under this specification:

* **Factor accumulation** is captured by  
  \[
  F \;=\; k^{\alpha}\,h^{1-\alpha},
  \]
  which aggregates the contributions of physical and human capital.

* **Relative productivity** is measured by  
  \[
  A \;=\;\frac{y}{F},
  \]
  isolating the portion of income differences that cannot be explained by factor inputs alone.  

**Calculation**
\vspace{-10pt}
```{r q5_data}
alpha <- 1/3          # capital share 

countries <- tribble(             # input given data 
  ~Country,      ~y,   ~k,   ~h,
  "Netherlands", 0.78, 1.12, 0.90,
  "Paraguay",    0.16, 0.15, 0.70,
  "Pakistan",    0.15, 0.08, 0.47
) %>% 
  mutate(                                   # desired calculations 
    Factor_Accum = k^alpha * h^(1 - alpha),
    Productivity = y / Factor_Accum,
    Factor_Share = 100 * Factor_Accum / y,
    Prod_Share   = 100 * Productivity / y 
  )
```

```{r q5_table, echo=FALSE}
knitr::kable(             # table output 
  countries %>%
    select(Country, y, k, h,
           `Factor Accum.` = Factor_Accum,
           Productivity,
           `Factor Share (%)`  = Factor_Share,
           `Prod. Share (%)`  = Prod_Share),
  digits = 2,
  caption = "Question 5 – Decomposing the 2017 income gap relative to the United States"
)
```

**Interpretation**  
Factor accumulation plays the largest role in explaining income relative to the United States in Paraguay; productivity plays the largest role in explaining income relative to the United States in the Pakistan.

---

### Question 6 – Growth Accounting for Argentina, Uruguay & Panama (1975‑2017)

**Method and Formulae**  
We assume a Cobb–Douglas production function in per-worker terms,  

\[
y = A\,k^{\alpha}h^{1-\alpha},
\]

with a typical capital share \(\alpha \approx \tfrac13\).

Log-differentiating yields  

\[
\hat{y} \;=\; \hat{A} \;+\; \alpha\,\hat{k} \;+\; (1-\alpha)\,\hat{h},
\]

where

* \(\hat{y}\) = growth of output per worker,  
* \(\hat{k}\) = growth of physical capital per worker,  
* \(\hat{h}\) = growth of human capital per worker,  
* \(\hat{A}\) = total factor productivity (TFP) growth (the Solow residual),  

so that we can derive

* **TFP growth:** \(\hat{A} = \hat{y} - \bigl[\alpha\,\hat{k} + (1-\alpha)\,\hat{h}\bigr]\)  
* **Factor accumulation growth:** \(\hat{FA} = \alpha\,\hat{k} + (1-\alpha)\,\hat{h}\).

**Calculation**
```{r q6_data}
tbl <- data.frame(                                 # Input given data
  Country   = c("Argentina", "Uruguay", "Panama"),
  g_y       = c(0.31, 1.46, 1.59),   # Growth of output per worker (%)
  g_k       = c(0.88, 1.59, 1.97),   # Growth of capital per worker (%)
  g_h       = c(0.79, 0.73, 0.84)    # Growth of human capital per worker (%)
)

alpha <- 1/3          # capital share

tbl <- tbl |>
  mutate(             # desired calculations
    g_FA = alpha * g_k + (1 - alpha) * g_h, # Factor accumulaton growth
    g_A = g_y - g_FA,                       # TFP growth
    share_FA = 100 * g_FA / g_y, # % of growth due to FA
    share_A = 100 * g_A / g_y  # % of growth due to TFP
)
```

```{r q6_table, echo=FALSE}
knitr::kable(
  tbl |>
    select(Country, g_y, g_k, g_h, g_FA, g_A, share_FA, share_A),
    digits = 2,
    col.names = c("Country",
                  "\\protect$\\hat{y}$ (\\%)", 
                  "\\protect$\\hat{k}$ (\\%)", 
                  "\\protect$\\hat{h}$ (\\%)", 
                  "\\protect$\\hat{FA}$ (\\%)", 
                  "\\protect$\\hat{A}$ (\\%)",
                  "FA Share of Growth (\\%)",
                  "TFP Share of Growth (\\%)"),
    caption = "Question 6 – Growth rate decomposition, 1975‑2017",
  format = "latex",
  booktabs = TRUE,
  escape = FALSE
  ) |>
  kable_styling(latex_options = "HOLD_position")
```

**Interpretation**  
In Argentina, factor accumulation accounts for the largest share of growth contribution among the three countries at 264.52% of growth; in Uruguay, total factor productivity accounts for the largest share of growth contribution among the three countries at 30.37%.

---

# Chapter 8: The Role of Technology in Growth

### Question 1 – The Rivalry and Excludability of Goods

| Good | Rival? | Excludable? |
|------|--------|-------------|
| National defense | No | No |
| A cookie | Yes | Yes |
| Password‑protected website | No | Yes |
| Fruit on a public square tree | Yes | No |

---

### Question 2 – Evaluating Patent Protection for Life Saving Drugs

**Advantages of patents**  

* Monopoly rents raise expected returns and stimulate risky, long-term R&D.  
* Patent disclosure places technical details in the public domain once exclusivity ends.  
* Clear intellectual property rights facilitate venture financing, licensing, and strategic alliances.  
* Temporary exclusivity aligns with regulatory approval timelines, helping recoup clinical-trial costs.  


**Disadvantages of patents**  

* Deadweight loss from monopoly pricing; patients in low income groups may be priced out.  
* Follow-on innovation can be blocked by “patent thickets” and evergreening strategies.  
* High prices distort R&D toward diseases with lucrative markets, neglecting rare or tropical illnesses.  
* Administrative, monitoring, and litigation costs consume public and private resources.  

**Possible alternative:** *Prize funds* (or advance market commitments) pay inventors for reaching pre‑specified health targets, after which the technology enters the public domain.  
*Advantages:* keeps the incentive to innovate while allowing competitive generic supply immediately; can be targeted to neglected diseases; reduces IP-related litigation.  
*Disadvantages:* requires large upfront public financing and careful contract design; risks mispricing the reward; demands robust verification of performance; may dampen incremental proprietary improvements without supplementary incentives.  

---

### Question 3 – Temporary Increase in \(\gamma_A\) in the One‑Country Model

```{r q3_setup, message=FALSE, include=FALSE}
library(tidyverse)

# Parameters
T          <- 60                 # periods
gamma_base <- 0.05               # baseline R&D share
gamma_high <- 0.10               # temporary high R&D share
t_start    <- 10                 # shock starts
t_end      <- 20                 # shock ends
mu         <- 10                 # idea price
A0         <- 1                  # initial productivity

# gamma_A path
gammaA <- rep(gamma_base, T)
gammaA[(t_start+1):(t_end)] <- gamma_high      # R uses 1-indexing

# Growth rates of A
gA <- gammaA/mu                 #  \hat A_t

# Simulate paths
A  <- numeric(T); y  <- numeric(T)
A[1] <- A0; y[1] <- A0*(1-gammaA[1])

for(t in 2:T){
  A[t] <- A[t-1]*(1+gA[t-1])
  y[t] <- A[t]*(1-gammaA[t])
}

baseline_A <- cumprod(c(A0, rep(1+gamma_base/mu, T-1)))[1:T]
baseline_y <- baseline_A*(1-gamma_base)

sim_df <- tibble(
  period = 0:(T-1),
  A_shock = A,
  A_base  = baseline_A,
  y_shock = y,
  y_base  = baseline_y
) %>% pivot_longer(-period,
                   names_to = c("series","scenario"),
                   names_sep = "_",
                   values_to = "value")
```

```{r q3_graphs, echo=FALSE}
ggplot(sim_df %>% filter(series=="A"),
aes(period, value, colour = scenario, linetype = scenario)) +
geom_line(size = 1) +
labs(y = "Productivity A (ratio scale)",
x = "Time",
title = "Path of Productivity A") +
scale_y_log10() +
theme_minimal()

ggplot(sim_df %>% filter(series=="y"),
aes(period, value, colour = scenario, linetype = scenario)) +
geom_line(size = 1) +
labs(y = "Output per worker y (ratio scale)",
x = "Time",
title = "Path of Output per Worker y") +
scale_y_log10() +
theme_minimal()
```

---

### Question 4 – Quantifying the ‘Catch‑Up’ Time after Raising \(\gamma_A\)

Parameters: \(L=1,\;\mu=5\).  Initially \(\gamma_A=0.5\) then *permanently* raised to \(0.75\).

**Model facts from §8.3**

* Per-worker output: \(y = A\,(1-\gamma_A)\).
* With a constant \(\gamma_A\), technology grows at \(\hat{A} = \tfrac{\gamma_A}{\mu}\).
* Hence \(\hat{y} = \tfrac{\gamma_A}{\mu}\) in the balanced growth path.

**Step 1 – Initial growth rate**

With \(L=1\), \(\mu = 5\), and \(\gamma_A = 0.5\):
\[
  \hat{y}_{\text{old}} = \frac{0.5}{5} = 0.10 \;\Rightarrow\; 10\%\text{ per year.}
\]

**Step 2 – Raise \(\gamma_A\) to 0.75**

* *Impact effect on output per worker:* \(y_{\text{old}} = A(1 - \gamma) = A(1 - 0.5) = A(0.5)\\ y_{\text{new}} = A(1 - \gamma) = A(1 - 0.75) = A(0.25)\)
, a 50 % drop.
* *New growth rate:* \(\hat{y}_{\text{new}} = \tfrac{0.75}{5} = 0.15\) $\Rightarrow$ 15 % per year.

The higher research share therefore produces a discrete fall in output but accelerates growth.

**Step 3 – Catch-up time**

Let \(t\) be the number of years after the policy change.  The new path overtakes the old one when
\[
   \hat{y}_{\text{new}}\,(1+g)^t =  \hat{y}_{\text{old}}
\]
Substituting given values,
\[
  A(0.25)(1+0.15)^t = A(0.50).
\]
Solving for $t$,
\[
  t\,\textup{ln}(1.15) = \textup{ln}(2).
\]

```{r q4_numeric}
mu        <- 5
gamma_old <- 0.50
gamma_new <- 0.25

# (i) Growth rate of productivity and output per worker (equal to gamma_A / mu)
g_A <- 0.15      # 15 % per year

# (ii) Time until the new path catches up with the old path (discrete compounding)
catch_up_years <- log(gamma_old / gamma_new) /
                  log(1 + g_A)

round(c(
        g_A = g_A,
        catch_up_years = catch_up_years), 3)
```

Output per worker **falls by one-half on impact** because the share of labor diverted to R&D rises, yet it subsequently grows faster. With technology growing at 15 % instead of 10 %, the economy needs roughly **5 years** for the high-R&D path to catch up to (and then exceed) the old trajectory.

---

### Question 5 – Two‑Country Model with Rise in $\gamma_{A,1}$

```{r q5_graph_2, echo=FALSE}

## 1.  Data ---------------------------------------------------------
df <- tibble(A_ratio = seq(1, 4, length.out = 300)) |>
  mutate(                                # single convex curve
    gA = 0.06 + 0.18 * (A_ratio - 1) + 0.025 * (A_ratio - 1)^2
  )

steady_old <- 0.32          # original steady-state growth rate  Â₁
steady_new <- steady_old + 0.10   # higher steady-state level    Â₂

# x-coordinates where the curve meets each horizontal line
old_ss <- with(df, approx(gA, A_ratio, xout = steady_old)$y)
new_ss <- with(df, approx(gA, A_ratio, xout = steady_new)$y)

y_max <- max(df$gA)

## 2.  Plot ---------------------------------------------------------
ggplot(df, aes(A_ratio)) +
  labs(title = "Country 1") +
  # Technology-growth locus (static)
  geom_line(aes(y = gA), colour = "#002F6C", linewidth = 1.3) +

  # Old and new steady-state lines
  geom_hline(yintercept = steady_old, colour = "#7AA0D2", linewidth = 0.8) +
  geom_hline(yintercept = steady_new, colour = "#7AA0D2", linewidth = 0.8) +

  # Dashed verticals to each intersection
  geom_vline(xintercept = old_ss, linetype = "dashed") +
  geom_vline(xintercept = new_ss, linetype = "dashed") +

  # Arrow showing the upward shift
  annotate("segment",
           x = new_ss + 0.25, xend = new_ss + 0.25,
           y = steady_old + 0.01, yend = steady_new - 0.01,
           arrow = arrow(length = unit(0.18, "cm")),
           colour = "#2a8d6f") +

  # Labels
  annotate("text", x = 3.8, y = df$gA[nrow(df)] + 0.01,
           label = "hat(A)[2]", parse = TRUE,
           hjust = 1, size = 4.2, colour = "#002F6C") +
  annotate("text", x = 3.8, y = steady_old - 0.05,
           label = "hat(A)[1]", parse = TRUE,
           hjust = 1, size = 4.2, colour = "#7AA0D2") +
  annotate("text", x = 3.8, y = steady_new + 0.05,
           label = as.character(TeX("$\\hat{A}_{1}'$")), parse = TRUE,
           hjust = 1, size = 4.2, colour = "#7AA0D2") +
  annotate("text", x = old_ss - .25, y = steady_old - 0.10,
           label = "Old steady\nstate", vjust = 1.1, size = 3.8) +
  annotate("text", x = new_ss + .25, y = steady_new + 0.10,
           label = "New steady\nstate", vjust = -0.1, size = 3.8) +

  # Axes
  scale_x_continuous(name = expression(A[1] / A[2]), limits = c(1, 4)) +
  scale_y_continuous(
    name   = TeX("Growth rate of technology, $\\hat{A}$"),
    limits = c(0, y_max + 0.05), expand = c(0, 0)
  ) +

  # Theme tweaks
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    axis.title.y     = element_text(margin = margin(r = 10))
  )
```

```{r q5_graph_1, echo=FALSE}
## 1.  Build the data frame ----------------------------------------------------
# X-axis runs from 1 (normalised) out to, say, 4
df <- tibble(A_ratio = seq(1, 4, length.out = 300)) |>
  mutate(
    # Original locus (light blue): a gently convex curve
    gA1 = 0.06 + 0.18 * (A_ratio - 1) + 0.025 * (A_ratio - 1)^2,

    # Up-shifted locus (dark blue): parallel upward move
    gA2 = gA1 + 0.10
  )

# Choose a steady-state growth rate (horizontal line)
steady <- 0.32
steady_old <- 0.32
steady_new <- steady_old + .10

# Locate where each curve crosses the steady-state line
old_ss <- with(df, approx(gA1, A_ratio, xout = steady)$y)  # old
new_ss <- with(df, approx(gA2, A_ratio, xout = steady)$y)  # new

# --- Arrow coordinates: between gA1 and gA2, but above the steady line -------
x_arrow <- new_ss + 0.25
g1x <- with(df, approx(A_ratio, gA1, xout = x_arrow)$y)
g2x <- with(df, approx(A_ratio, gA2, xout = x_arrow)$y)
pad <- 0.01
y0 <- g1x + pad  # start just above the straight line (and gA1)
y1 <- g2x - pad                     # end just below gA2

## 2.  Draw the figure ---------------------------------------------------------
ggplot(df, aes(A_ratio)) + 
  labs(title = "Country 2") +
  geom_line(aes(y = gA1), colour = "#002F6C", linewidth = 1.3) +
  geom_line(aes(y = gA2), colour = "#002F6C", linewidth = 1.3) +
  geom_hline(yintercept = steady, colour = "#7AA0D2", linewidth = 0.7) +
  geom_vline(xintercept = old_ss, linetype = "dashed") +
  geom_vline(xintercept = new_ss, linetype = "dashed") +
  annotate(
    "segment",
    x = x_arrow, xend = x_arrow,
    y = y0,     yend = y1,
    arrow = arrow(length = unit(0.18, "cm")),
    colour = "#2a8d6f"
  ) +
  annotate("text", x = old_ss + .25, y = steady - 0.05,
           label = "Old steady\nstate", vjust = 1.1, size = 3.8) +
  annotate("text", x = new_ss - .25, y = steady + 0.15,
           label = "New steady\nstate", vjust = 1.1, size = 3.8) +
  annotate("text", x = 3.8, y = df$gA1[nrow(df)] - 0.15,
         label = "hat(A)[2]",           # <-- character
         parse = TRUE,                  # <-- tell ggplot to parse it
         hjust = 1, size = 4.2,
         colour = "#002F6C") +
  annotate( "text",
            x     = 3.8,
            y     = df$gA2[nrow(df)] + 0.01,
            label = as.character(TeX("$\\hat{A}_{2}'$")),
            parse = TRUE,
            hjust = 1, 
            size  = 4.2,
            colour = "#002F6C") +
  annotate("text", x = 3.8, y = steady_old - 0.05,
         label = "hat(A)[1]",           # <-- character
         parse = TRUE,
         hjust = 1, size = 4.2,
         colour = "#7AA0D2") +
  scale_x_continuous(name = TeX("$A_1/A_2$"), limits = c(1, 4)) +
  scale_y_continuous(
    name   = TeX("Growth rate of technology, $\\hat{A}$"),
    limits = c(0, max(df$gA2) + 0.05),
    expand = c(0, 0)
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    axis.title.y     = element_text(margin = margin(r = 10))
  )

```

---

*End of Assignment*
